\vfill World of Warcraft's New Way to Get You Hooked

### 2013-04-01 paper.tex
"Final" version

% Cover page with abstract: A cover page included: title, name, date, "CSC 300," and an abstract. The abstract is one or two sentences of context about your issue.  Provide your focus question. Mention one or two relevant arguments.  Briefly explain your conclusion and why it is proper ethically.

@@

% [5%] Abstract (75-100 Words Max)
% __ One or two sentences of context about your issue.
% __ Provide your focus question.
% __ Mention one or two relevant arguments.
% __ Briefly explain your conclusion and why it is proper ethically. (Optimal total is 7 sentences or so...)

@@

%start 2 column format
\begin{multicols}{2}
%Start numbering first page of content as page 1
\setcounter{page}{1}
%%%%%%%%%%%%%%%%%%%%
%%% Known Facts  %%%
%%%%%%%%%%%%%%%%%%%%

% [10%] Facts (500 Words Max) This is not a general "introduction" - it has a very specific purpose.
%__ Give general context for the issue.
%__ This section must present *ONLY ENOUGH* substantiated facts to raise the question for the reader (not to answer the question, additional facts may be introduced during analysis).
% __ The facts section should end up leading the reader to ask the question you are about to ask (a controversy or a case ending poorly is a great way to do this).
% __ Should not indicate anything about your answer to the question or hint at any particular conclusion.
% __ Do not include the definitions of ethical systems or laws as background. Just raise your focus question.
\section{Facts}

Blizzard Entertainment's World of Warcraft is an incredibly popular game with over 10 million subscribers. \cite{WoWPlayerCount} Players of World of Warcraft spend an average of 22 hours per week playing. \cite{PlayerMotivations} World of Warcraft is free to play provided the player is willing to accept some restrictions. None of those restrictions hinder the number of hours one can play. \cite{StarterEditionFAQ} Multiple studies have shown that time spent playing a MMORPG, like World of Warcraft, correlates to addiction level in some individuals. \cite{ProblemUsageAmongHighlyEngagedPlayers}\cite{JustAComputerGameRight} The intensity with which many players play World of Warcraft or other MMORPGs has led to deaths. \cite{SnowlyDeath}\cite{ShawnDeath}

@@

% [5%] Focus Question (One Non-Compound Sentence Max followed by a short paragraph of 50 words Max)
% __ Must be focused on a particular, specific case. More detail / less abstract is better.
% __ Must call for a binary answer (yes / no) that you will draw based on your research and reasoning.
%__ Should be followed (separately) by a single paragraph of 50 words max explaining the importance of this question and its relevance to software engineers (why should we care?). There should be a citation to an external source to validate such importance.

\section{Research Question}
Is Blizzard's use of World of Warcraft: Starter Edition ethical? \\

World of Warcraft: Starter Edition could precipitate an ``outbreak'' of videogame addiction leading to terrible incidents. \cite{SnowlyDeath} Deaths and addiction linked to World of Warcraft would also cause a sharp decline in the sale of games. Blizzard made \$4.85 billion last year and has around 4700 employees. \cite{BlizzardFinances} \cite{BlizzardEmployeeCount} At its worst the company could fail and those 4700 employees would be without income and looking for work.

@@

% [10%] Others' Arguments (500 Words Max)
% __ Summarize the main arguments others have made, citing sources for each one. Make each summary under a separate, descriptive subheading.
% __ Must cover the strongest arguments on both sides of the issue. Make them clear, concise and objective (no opinion as to their strength).
% __ Do not use bare ethical principles (including the Code) in this section - they are not "other's" arguments about your focus question.
% __ Summarize all arguments very briefly at the end of this section.

\section{Arguments}
\subsection{Arguments Affirmative}
\subsubsection{Differing Player Archetypes}
Users play MMORPGs for a variety of reasons. Sweeping generalizations are often made without basis by grouping all players into a single type that is prone to addiction. \cite{PlayerMotivations}
\subsubsection{More Time Online Does Not Mean Addiction}
Many people who are characterized as spending ``excessive'' (not defined in the article) time online do not show symptoms of addiction. These individuals show signs of psychological fulfillment. Often these people would be unable to find such fulfillment elsewhere due to disabilities, circumstances, or social discomfort. Some people who are isolated or autistic find comfort in the use of the internet and games to create social relationships they would not have cultivated otherwise. \cite{IsThereEvidenceOfInternetAddiction}
\subsubsection{Excessive Users Do Not Exhibit Addiction Symptoms}
Users that spend more time than average playing the game exhibit social fulfillment. Many users who play more than average are playing MMORPGs for social aspects of the in-game world. These users would just as likely find fun elsewhere. \cite{ExcessiveUseForSocialAspects}
\subsection{Arguments Negative}
\subsubsection{More Intense Play By A Younger Audience Leads to Addiction}
``MMORPGs are played more intensely than video games,'' and are played primarily by a younger audience. Persons under the age of 27, the majority of players of MMORPGs, are far more susceptible to addiction due to a lack of self identity and self confidence. \cite{YoungerAudiencesGetAddicted}
\subsubsection{Increased Time Playing Leads to Life Problems}
The longer a person plays World of Warcraft, the more likely they are to experience issues in their lives. \cite{ProblemUsageAmongHighlyEngagedPlayers}
\subsubsection{Addictive Games Have a Social Responsibility}
Current efforts to warn players of online games' addictiveness are minimal and often non-existent. Warning labels outlining the possibility of addition should be printed on game boxes and companies should contact users who play in excess of a set threshold. \cite{MMORPGSocialResponsibility}\\

@@

% [70%] Your Analysis (About 3000 Words)
% __ Start with a paragraph showing why the SE Code applies to your focus question. Do not merely conclude, but show the reader by application of Code principles.
% __ Liberally utilize sub-headings organized by the relevant SE Code sections and particular arguments to delineate your lines of reasoning.
% __ All your analysis must be thoroughly supported by reason and logic.
% __ All claims must be supported by reputable primary sources and formal data.
% __ Have 2-4 distinct sections of the SE code utilized in your analysis.
%__ Code section 1 can be central to one piece of analysis only. For example, if you determine section 1.03 applies, then no other section 1 Code tenet may be used later as a section heading.
% __ If discussion about "public good" is used, there must be data to support it. Simply writing "it benefits the general public because it would make many people happy" is insufficient.
% __ Utilitarian, Rawlsian and Deontological analyses must be integrated into your analysis where appropriate. They should support your SE Code analysis, not be isolated from it. (Exceptions must be approved by Dr. Turner.)
% __ Class readings should be referenced as appropriate to support your Code analysis.
% __ There should be a clear cohesiveness to the analysis such that each argument logically flows into the next and the reader is logically convinced that your conclusion has strength. Your writing must explicitly provide substantial arguments through logic and data that will gain the respect of a skeptic.
% __ Do not define ethical systems in your analysis unless they are required and are very brief (a few sentences maximum).
% __ Opinions + Assumptions + Feelings > /dev/null

@@

\subsection{Definition of ``addiction''}
Because this paper will often reference psychological addiction, it is important to make sure this notion is concrete before continuing.\\
A person who is addicted will pathologically pursue rewards and/or relief through substance abuse or other behaviors, in this case playing World of Warcraft. \cite{DefinitionOfAddiction}

@@

\subsection{SE Code Section 1.02\\Moderate the interests of the software engineer, the employer, the client and the users with the public good.}
In the next few sections this tenet is broken down into its parts. I will explore the ``interests of the software engineer,'' in this case World of Warcraft's software engineers, the ``interests of ... the employer,'' in this case Blizzard Entertainment, and lastly the ``interests of ... the client and the users,'' the player. Following that will be a discussion of the ``public good'' using Utilitarianism to guide my definition.\\
``World of Warcraft's Engineers shall moderate the interests of themselves, Blizzard Entertainment, and the players of World of Warcraft with the public good.''
\footnote{Ryan did a great job with his comments on the actual document.}
\subsubsection{Interests of the Software Engineer}
It is common sense that a software engineer strives to make working software. For World of Warcraft's software engineers, this success is measured by how many players enjoy new content and continue to play the game. Considerations are made for balance and gameplay as well as story. \cite{DevWatercooler} These engineers are also interested in keeping their jobs and continuing to make money. In order to keep their jobs, their interests must lie in getting players to come back to the game world often and play the game.
\subsubsection{Interests of Blizzard Entertainment}
Blizzard Entertainment is a publicly traded company. \cite{BlizzStock} As such its main goal is to show its shareholders that it is profiting and increases the volume of business it handles. In no uncertain terms, Blizzard Entertainment wants as many people to buy and play their games, World of Warcraft included, as possible. Since World of Warcraft is subscription-based \cite{WoWSubscription}, Blizzard Entertainment earns money every single month from World of Warcraft players. None of the other games that Blizzard makes are subscription based. \cite{BlizzGameList} Blizzard Entertainment's interests are in getting more players to buy their games, particularly World of Warcraft which nets them money every month.
\subsubsection{Interests of the Player}
By purchasing a game like World of Warcraft, players are committing themselves to pay Blizzard Entertainment every month. Some players are reaping benefits that they would not be able to get otherwise, such as an increase in social interactions. \cite{IsThereEvidenceOfInternetAddiction} One particular study found that MMORPG users play for a mixture of ten reasons which break down into three categories: achievement, social, and immersion. \cite{PlayerMotivations} (All information following in this section is from the same source.) MMORPGs offer a sense of achievement through completing quests, challenging others, and optimizing character performance. MMORPGs offer the chance to socialize with people you've never met, join groups to accomplish tasks, and form relationships with people through guilds and group events. MMORPGs also offer a sense of immersion through their expansive worlds that can be explored, the opportunity to role-play with others, and the ability to customize your characters in a nearly infinite set of ways. The interests of the player boil down to achievements, social interaction, and immersion in a new world.
\subsubsection{The ``Public Good''}
I define the ``public good'' to be a population that is not addicted to World of Warcraft and therefore a population that exhibits the ability to abstain from World of Warcraft, have unimpaired behavior control, and not crave playing World of Warcraft.

In order to argue that the ``public good'' is being upheld I assert that Utilitarianism is the most effective ethical system to use because it judges actions based on the net happiness of everyone. Utilitarianism is ``[t]he creed which ... holds that actions are right in proportion as they tend to promote happiness, wrong as they tend to produce the reverse of happiness.'' \cite{Utilitarianism}

World of Warcraft has over 10 million subscribers \cite{WoWPlayerCount} who average 22 hours of play time per week. \cite{PlayerMotivations} For those users that spend above average time, they are most likely just enjoying the social aspects of the game. \cite{ExcessiveUseForSocialAspects} These people would have fun doing other activities instead and would not be bothered by not being able to play the game for a day or two. \cite{ExcessiveUseForSocialAspects} PSYCHOLOGICAL FULLFILLMENT, OTHER ARGUMENTS, READ YOUR DAMN PAPERS AND GET MORE FACTS

The most shocking statistic came from Dr. Maressa Orzack, founder and coordinator of Computer Addiction Services \cite{CompAddictionServices}, who cites that up to 40\% of World of Warcraft players could be addicted. \cite{FortyPercentAddicted} If everyone in the world played World of Warcraft, a total of over 7 billion people \cite{WorldPopulation}, and Dr. Ozack's prediction is correct, 2.8 billion people would be addicted. Based purely on the numbers that leaves 60\% of the world population who are not addicted.

Having a majority of players who are \emph{not} addicted, and thus playing for pleasure or other purposes, is in line with the Utilitarian view of ethical and therefore in line with the ``public good.''

\subsubsection{Ethical Conclusion}
In order for Blizzard's actions to be ethical, its engineers must ``moderate the interests of themselves, Blizzard Entertainment, and the players of World of Warcraft with the public good.''
The public good is being upheld, Blizzard Entertainment's actions are ethical in accordance with SE Code section 1.02.

\begin{comment}
\subsection{SE Code Section 2.07\\Identify, document, and report significant issues of social concern, of which they are aware, in software or related documents, to the employer or the client.}
DICE Interview? Gamastura, sounded concerned.

Opponents have suggested warning labels and/or contacting players who exhibit signs of addiction. I may have to concede this point, I see nothing wrong with a few added measures.
``Tip: Remember to take all things in moderation (even World of Warcraft!)'' - In-Game Tip! WoW is at least putting some effort forth.

\end{comment}

>>>>

@@

% [5%] Abstract (75-100 Words Max)

@@

%start 2 column format
\begin{multicols}{2}
%Start numbering first page of content as page 1
\setcounter{page}{1}

%%%%%%%%%%%%%%%%%%%%
%%% Known Facts  %%%
%%%%%%%%%%%%%%%%%%%%

\section{Facts}

Blizzard Entertainment's World of Warcraft is an incredibly popular game with over 10 million subscribers. \cite{WoWPlayerCount} Players of World of Warcraft spend an average of 22 hours per week playing. \cite{PlayerMotivations} World of Warcraft is free to play provided the player is willing to accept some restrictions. None of those restrictions hinder the number of hours one can play. \cite{StarterEditionFAQ} Multiple studies have shown that time spent playing a MMORPG, like World of Warcraft, correlates to addiction level in some individuals. \cite{ProblemUsageAmongHighlyEngagedPlayers}\cite{JustAComputerGameRight} The intensity with which many players play World of Warcraft or other MMORPGs has led to multiple deaths. \cite{SnowlyDeath}\cite{ShawnDeath}\cite{ParentsBlameWoWForDeath}\cite{DeathAfter50Hours}\cite{YouStoleMySwordNowYouDie}

@@

\section{Focus Question}
Is Blizzard's use of World of Warcraft: Starter Edition ethical? \\

World of Warcraft: Starter Edition could precipitate an ``outbreak'' of videogame addiction leading to terrible incidents. \cite{SnowlyDeath}\cite{ShawnDeath}\cite{ParentsBlameWoWForDeath}\cite{DeathAfter50Hours}\cite{YouStoleMySwordNowYouDie} Deaths and addiction linked to World of Warcraft would cause a sharp decline in the sale of games. Blizzard made \$4.85 billion last year and has around 4700 employees. \cite{BlizzardFinances}\cite{BlizzardEmployeeCount} At its worst the company could fail and those employees would be without income and looking for work.

@@

\section{Arguments}
\subsection{Arguments Affirmative}
\subsubsection{Differing Player Archetypes}
Users play MMORPGs for a variety of reasons. Sweeping generalizations are often made without basis by grouping all players into a single type that is prone to addiction. \cite{PlayerMotivations}
\subsubsection{More Time Online Does Not Mean Addiction}
Many people who are characterized as spending ``excessive'' (not defined in the article) time online do not show symptoms of addiction. These individuals show signs of psychological fulfillment. Often these people would be unable to find such fulfillment elsewhere due to disabilities, circumstances, or social discomfort. Some people who are isolated or autistic find comfort in the use of the internet and games to create social relationships they would not have cultivated otherwise. \cite{IsThereEvidenceOfInternetAddiction}
\subsubsection{Excessive Users Do Not Exhibit Addiction Symptoms}
Users that spend more time than average playing the game exhibit social fulfillment. Many users who play more than average are playing MMORPGs for social aspects of the in-game world. These users would just as likely find fun elsewhere. \cite{ExcessiveUseForSocialAspects}
\subsection{Arguments Negative}
\subsubsection{More Intense Play By A Younger Audience Leads to Addiction}
``MMORPGs are played more intensely than [offline] video games,'' and are played primarily by a younger audience. Persons under the age of 27, the majority of players of MMORPGs, are far more susceptible to addiction due to a lack of self identity and self confidence. \cite{YoungerAudiencesGetAddicted}
\subsubsection{Increased Time Playing Leads to Life Problems}
The longer a person plays World of Warcraft, the more likely they are to experience issues in their lives. \cite{ProblemUsageAmongHighlyEngagedPlayers}
\subsubsection{Addictive Games Have a Social Responsibility}
Current efforts to warn players of online games' addictiveness are minimal and often non-existent. Warning labels outlining the possibility of addition should be printed on game boxes and companies should contact users who play in excess of a set threshold. \cite{MMORPGSocialResponsibility}\\

@@

@@

\subsection{Definition of ``addiction''}
Because this paper will often reference psychological addiction, it is important to make sure this notion is concrete before continuing.\\
A person who is addicted to something will pathologically pursue rewards and/or relief through substance abuse or other behaviors, in this case playing World of Warcraft. \cite{DefinitionOfAddiction}

@@

\subsection{How Long Does It Take To Hit The Cap?}
With World of Warcraft's Starter Edition, the most debilitating restriction is the level cap. Starter Edition accounts are capped at level 20. \cite{StarterEditionFAQ} With the current highest level at 90, the cap happens quite early in gameplay. I created two different characters and took them from level 1 to 20. It took me 6 and 8 hours respectively to reach the cap.\footnote{I am an expert player with over two months of in-game time logged on my account. A newer player would likely take longer than this to reach level 20.}
Some users believe they have become addicted to MMORPGs from the moment they first play them. \cite{JustAComputerGameRight} There appears to be no cut and dry length of time that it takes to become addicted to something, different people become addicted to things at different rates.
Because there's no evidence that time is a factor in becoming addicted, the real argument becomes whether or not most users become addicted at all, whether the dangers of becoming addicted are outweighed by other factors, and whether Blizzard is doing it's part to warn its users.

\subsection{SE Code Section 1.02\\Moderate the interests of the software engineer, the employer, the client and the users with the public good.}
In the next few sections this tenet is broken down into its parts. I will explore the ``interests of the software engineer,'' in this case World of Warcraft's software engineers, the ``interests of ... the employer,'' in this case Blizzard Entertainment, and lastly the ``interests of ... the client and the users,'' the player. Following that will be a discussion of the ``public good.''\\
In this situation SE Code Section 1.02 becomes: ``World of Warcraft's Engineers shall moderate the interests of themselves, Blizzard Entertainment, and the players of World of Warcraft with the public good.''
\subsubsection{Interests of the Software Engineer}
It is common sense that a software engineer strives to make working software. For World of Warcraft's software engineers, this success is measured by how many players enjoy new content and continue to play the game. Considerations are made for balance and gameplay as well as story. \cite{DevWatercooler} These engineers are also interested in keeping their jobs and continuing to make money. In order to keep their jobs, their interests must lie in getting players buy the software and then to come back to the game world often and play the game.
\subsubsection{Interests of Blizzard Entertainment}
Blizzard Entertainment is a publicly traded company. \cite{BlizzStock} As such its main goal is to show its shareholders that it is profiting and increasing the volume of business it handles. In no uncertain terms, Blizzard Entertainment wants as many people to buy and play their games, World of Warcraft included, as possible. Since World of Warcraft is subscription-based \cite{WoWSubscription}, Blizzard Entertainment earns money every single month from World of Warcraft players. None of the other games that Blizzard makes are subscription based. \cite{BlizzGameList} That means that Blizzard only ever receives one payment for any of its other games from customers. Blizzard Entertainment's interests are in getting more players to buy their games, particularly World of Warcraft which nets them money every month in addition to the initial cost.
\subsubsection{Interests of the Player}
By purchasing a game like World of Warcraft, players are committing themselves to pay Blizzard Entertainment every month. \cite{WoWSubscription} Some players are reaping benefits that they would not be able to get otherwise, such as an increase in social interactions. \cite{IsThereEvidenceOfInternetAddiction} Users who play the game rarely navigate the world alone, instead they form groups with other players, join guilds, and tackle quests and dungeons that require teamwork and talking to others.
One particular study found that MMORPG users play for a mixture of ten reasons: advancement, mechanics, competition, socializing, relationship, teamwork, discovery, role-playing, customization, and escapism. These ten reasons break down into three components: achievement, social, and immersion. \cite{PlayerMotivations} The achievement component encompasses achievement, ``the desire to gain power, progress rapidly, and accumulate in-game symbols of status or wealth,'' mechanics, ``having an interest in analyzing the underlying rules and system in order to optimize character performance,'' and competition, ``the desire to challenge and compete with others.'' The social component deals with socializing, ``having an interest in helping and chatting with other players,'' relationship, ``the desire to form long-term meaningful relationships with others,'' and teamwork, ``deriving satisfaction from being part of a group effort.'' Lastly, the immersion component is discovery, ``finding and knowing things that most other players don't know about,'' role-playing, ``creating a persona with a background story and interacting with other players to create improvised story,'' customization, ``having an interest in customizing the appearance of their character,'' and escapism, ``using the online environment to avoid thinking about real life problems.'' \cite{PlayerMotivations} (All information following in this section is from the same study. \cite{PlayerMotivations}) MMORPGs offer a sense of achievement through completing quests, challenging others, and optimizing character performance. MMORPGs offer the chance to socialize with people you've never met, join groups to accomplish tasks, and form relationships with people through guilds and group events. MMORPGs also offer a sense of immersion through their expansive worlds that can be explored, the opportunity to role-play with others, the ability to customize your characters in a nearly infinite set of ways, and the chance to step into a different world where your real-world problems fade away. The interests of the player boil down to achievements, social interaction, and immersion in a world outside their own.

\subsubsection{``Public Good'' Breakdown}
I define the public good to be the population's mental and physical welfare. A healthy population is not addicted to World of Warcraft and therefore exhibits the ability to abstain from World of Warcraft, have unimpaired behavior control, and not crave playing World of Warcraft.

I will take two approaches to arguing that the public good is being upheld. First will be a purely numerical approach to public good, `` ``Public Good'' By The Numbers'', followed by an argument that the public good is not only being upheld, but improved by World of Warcraft, `` ``Public Good'' Being Improved''.

\subsubsection{``Public Good'' By The Numbers}
In order to argue that the public good is being upheld I assert that Utilitarianism is the most effective ethical system to use because it judges actions based on the net happiness of everyone. Utilitarianism is ``[t]he creed which ... holds that actions are right in proportion as they tend to promote happiness, wrong as they tend to produce the reverse of happiness.'' \cite{Utilitarianism} A utilitarian view in this case would be that World of Warcraft does not create addiction in the majority of the world and therefore does not diminish mental or physical welfare of the public.

The most shocking statistic I found came from Dr. Maressa Orzack, founder and coordinator of Computer Addiction Services \cite{CompAddictionServices}, who cites that up to 40\% of World of Warcraft players could be addicted. \cite{FortyPercentAddicted} If everyone in the world played World of Warcraft, a total of over 7 billion people \cite{WorldPopulation}, and Dr. Ozark's prediction is correct, 2.8 billion people would be addicted. Based purely on the numbers that leaves 60\% of the world's population that is not addicted.

\subsubsection{``Public Good'' Being Improved}
World of Warcraft has over 10 million subscribers \cite{WoWPlayerCount} who average 22 hours of play time per week. \cite{PlayerMotivations} A study published in \emph{Cyberpsychology \& Behavior} \cite{ExcessiveUseForSocialAspects} states, ``[i]t is clear that MMORPG users have a tendency to spend many more hours devoted to their game and find the social aspects of the in-game world more pleasant and satisfying than what occurs in the real world. However, MMORPG users do not seek self-confidence in-game, would find fun elsewhere if MMORPGs did not exist, and would not feel irritated if they did not have the chance to play for 1 day. This would suggest that, as much as MMORPG users enjoy the time they spend in-game, even more so than real-life activities with friends, they are not addicted.'' World of Warcraft players are passionate about their experiences in the game, but they do not show signs of addiction.

In a case study of five ``problematic'' internet and game users, users who are seen by family and friends as having extreme computer addiction issues, only two showed signs of true addiction while the remaining three seemed to be using the internet to counter-act other deficiencies. \cite{IsThereEvidenceOfInternetAddiction} These three users who were not psychologically addicted used the internet and games to connect with others in ways they couldn't previously. Factors keeping them from doing so prior were due to physical appearance, social anxiety, and mental disability. One European man found a wife in the United States and moved to be with her, improving both their lives, thanks to his extended internet use. \cite{IsThereEvidenceOfInternetAddiction} These people use the internet and games as a way to enrich their lives, not detract from them.

During a panel at D.I.C.E. 2012 \cite{DICEInterview} that included Mike Epps, President of Epic Games, Ted Price, CEO and Founder of Insomniac Games, and Frank Pearce, current VP of product development and one of the three founders of Blizzard, Frank Pearce had this to say about World of Warcraft:
\begin{quote}
[W]e hear stories all the time, in fact right when we were walking out for the break, someone came up to me and said ``I just want you to know, I'm in this industry because I suffered a back injury and I spent six years playing Warcraft and decided I wanted to get into the games industry.'' We get letters from players all the time that can't experience the real world in the way that they would like to because of physical limitations and can live a completely free experience in a game like World of Warcraft.
\end{quote}
Blizzard Entertainment is doing so many people a service that they actually receive letters thanking them for making such an immersive experience.

In a three year study of over 30,000 MMORPG users, over 50\% of users said that they learned valuable leadership skills that have helped them outside of the game. \cite{MotivationsAndDerviedExperiences} To break it down further, 55.5\% of users believed they learned about conflict mediation, 48.4\% believed they learned about properly motivating a group, and 43.8\% believed they learned about persuading other members when a change in goals was necessary. \cite{MotivationsAndDerviedExperiences}
The study also showed that 45.8\% of users made friendships in the game that were comparable to or better than friendships they cultivated outside the game. 23.8\% of users had they most rewarding or satisfying experience in the game, as opposed to outside of the game, in the last 7 or 30 days. 28.6\% of users had their most annoying or infuriating experiences over the past 7 or 30 days occurring in the game. \cite{MotivationsAndDerviedExperiences}

Many features in World of Warcraft drive the players to socialize. \footnote{Having over two months of play time in-game with the last play session less than a week ago, I consider myself qualified to discuss the mechanics of World of Warcraft without citation.} The best gear and most prestigious titles are only available to users who group together to take on challenges that cannot be conquered by oneself. These challenges require intense coordination and communication as well as skilled players. Without grouping together, players have no chance of getting these items or attaining these statuses.
Guilds offer perks to their members through mechanics built into the game. As guilds grow in size and their members complete quests and dungeons, these guilds level up and offer rewards such as experience boosts allowing members to level faster, speed boosts allowing members to traverse the map faster, reputation gain boosts allowing members to increase their reputation with factions in the game that offer rewards, durability increases to its members' armor so that the armor does not need to be repaired nearly as often, increases in output from its members professions, and certain powerful spells that can be cast when guild members take on challenges together.

Friendships and powerful experiences help us grow as individuals. MMORPGs, like World of Warcraft, are providing their users with leadership skills, strong friendships, and powerful experiences.

\subsubsection{Ethical Conclusion}
In order for Blizzard's actions to be ethical, its engineers must ``moderate the interests of themselves, Blizzard Entertainment, and the players of World of Warcraft with the public good.''

The interests of the software engineer, creating content that players love, and the interests of the player are complimentary. Players are interested in great content and a full-featured experience and Blizzard's engineers are interested in creating that exact experience and content.
Blizzard Entertainment's interests of growing as a company and keeping its shareholders satisfied are only improved by happy players. Happy players will recommend Blizzard and its products to friends as well as continue to pay monthly fees for World of Warcraft and buy Blizzards other games.

Having a majority of players who are \emph{not} addicted, and thus playing for pleasure or other purposes, is in line with the Utilitarian view of ethical and therefore in line with a cold, calculated numerical view of ``public good.''

Not only are a majority of users not addicted, many of these users are cultivating life skills, strong friendships, and powerful emotional experiences. These users are growing as individuals and going out into the world as stronger people with leadership and teamwork skills.
Some users are even overcoming physical and mental disabilities through use of the game. These users would be mentally and physically distraught without World of Warcraft and thus the public good would be diminished. Because the game is helping these users from having poor welfare, World of Warcraft, and other MMORPGs, can be seen as \emph{improving} the public good.

Blizzard's software engineers are moderating all parties' interests. The public good is not only being upheld, but improved. Therefore, Blizzard Entertainment's actions are not only ethical in accordance with SE Code section 1.02, but can be seen as extraordinary, going beyond moderating their interests and improving the public good.

\subsection{SE Code Section 2.07\\Identify, document, and report significant issues of social concern, of which they are aware, in software or related documents, to the employer or the client.}
Software Engineers are required to make note of ``significant issues'' and make sure that users are aware of these issues. It's no stretch to consider psychological addiction to be a ``significant issue of social concern'' and therefore it is Blizzard Entertainment's responsibility to ``identify, document, and report'' this issue. As before, ``the client'' is the player and the ``software'' is World of Warcraft.

During a panel at D.I.C.E. 2012 \cite{DICEInterview} Frank Pearce answered a question about World of Warcraft addiction:
\begin{quote}
Mike Epps: It looks like I've put a good half man-year into one of my [World of Warcraft] characters. I'm not gonna say ``How do you sleep at night?'' ... But, do you think about that? The implications of this industry, of entertaining?\\
\\
Frank Pearce: Yeah, I mean we think about it all the time. ... If there's something that you're really really passionate about and you're consuming the media, whether it's videogames or a book or a television series... that means someone's created some great content, but I mean we have to consume it in moderation.
\end{quote}
Frank Pearce, and surely others at Blizzard, are aware of the issue of addiction, taking care of the clause ``of which they are aware.''

The code section in this case becomes ``World of Warcraft's software engineers shall identify, document, and report the issue of addiction to World of Warcraft to its players.''

World of Warcraft has a number of loading screens: when you log into the game, move across the map, and do a variety of other things.\footnote{Having over two months of play time in-game with the last play session less than a week ago, I consider myself an expert in the field of World of Warcraft, and thus needn't cite anyone but myself.} On each of these loading screens is a little segment of text with various ``tips'' that provide humor, help you in the game, or give you some advice. One such tip is: ``Tip: Remember to take all things in moderation (even World of Warcraft!)'' \cite{LoadingScreenTips} Blizzard's software engineers are specifically identifying and documenting the issue of addiction and suggesting time away from World of Warcraft.

In World of Warcraft one of the primary goals is to level up your character, unlocking new skills and becoming more powerful. In order to level up you complete quests and kill enemies, both of which give you experience points. When you collect enough experience points, you level up. If you log out of the game when standing in certain locations, usually at ``inns'' in big towns, you become ``rested'' for some period of time which increases the longer you are logged out. Being rested when you log back in, you gain twice as much experience from killing enemies as when you're not rested. This state of being ``rested'' wears off as you play and adventure in the world. In essence, logging out of the game actually helps you in the game.

Some experts have suggested warning labels for games like World of Warcraft as well as contacting players who exhibit signs of addiction, such as play sessions in excess of some period of time. \cite{MMORPGSocialResponsibility} I see nothing wrong with a few added measures that go hand in hand with the idea of identifying and reporting the issue of addiction.

Blizzard is clearly making the effort to get users to play in moderation. They even offer rewards for being offline. They are not using warning labels or contacting players, however.

\subsubsection{Ethical Conclusion}
Blizzard Entertainment and its software engineers have shown knowledge of the addiction issue and have identified and reported it in-game as well as implemented rewards that encourage off-line time.

Although Blizzard could be doing more, they have put in effort to help users and therefore they are ethical in identifying, documenting, and reporting the issue of addiction as well as going beyond the requirements and encouraging time away from the game.

\section{Conclusion}
There is no link between time spent doing an activity and addiction to it, for every person the timeline is different.

Blizzard Entertainment has moderated its interests, and documented the possibility of addiction. Blizzard has also gone above and beyond and improved the public's welfare while offering players rewards for \emph{not} playing.

World of Warcraft is a game that pushes its users to interact with each other and offers users the chance for achievement, socializing, and immersion in a world unlike their own. Through its mechanics and gameplay users learn valuable life and leadership skills and can interact with others to achieve a common goal.

Is Blizzard's use of World of Warcraft: Starter Edition ethical? Yes.
